Deriving Random Generators via\\ Higher-Level Representations

### 2019-01-17 paper/sections/hrep.tex
WIP

In this work we define type level combinators which let the user combine the
automatically derived |HRep|s in several ways.
%
In first place, we define a type combinator ($\_^\bullet$) to tag a |HRep| to be
terminal, i.e., a representation that is allowed be generated when the
generation size gets exausted:
%
\begin{code}
data (f_term) a = Term (f a)
\end{code}

@@

Unlike the other combinators, to randomly generate a fixed point of a certain
representation |f|, we do not parametrize the generation of the recursive fields
of the inner represention over an external generator |gen_r|.
%
Instead, we can also tie the recursive knot in our random generator by piping
the generation of the inner |HRep| parametrized recursively by the itself:

@@

@@

@@

\begin{align*}
  &\evalrep{|Mk_foo_1|\ x\ y}{Exp} = \\
  &\ \ |Add (Add eval_x_Exp (Val 50)) (Add (Val 25) eval_y_Exp)|\\
  &\evalrep{|Mk_foo_2|\ x\ y}{Exp} = \\
  &\ \ |Mul (Val 50) (Mul (Val x) eval_y_Exp)|
\end{align*}

@@

@@

@@

@@

@@

@@

@@

@@

This previous definition can be interpreted graphically as it is shown in the
Figure \ref{fig:hrep}.
%
Curly arrows represent the structural information extracted using
meta-programming.

@@

>>>>

In this work we define type level combinators which let the user combine the
automatically derived |HRep|s in several ways.
%
In first place, we define a type combinator (|term|) to tag a |HRep| to be
terminal, i.e., a representation that is allowed be generated when the
generation size gets exausted:
%
\begin{code}
data (f_term) a = Term (f a)
\end{code}

@@

Unlike the other combinators, to randomly generate a fixed point of a certain
representation |f|, we do not parametrize the generation of the recursive fields
of the inner represention over an external generator |gen_r|.
%
Instead, we can tie the recursive knot in our random generator by piping the
generation of the inner |HRep|, and calling itself to generate recursive values:

@@

So far we have introduced the representation machinery required to represent the
random generation of a target data type considering only the structure encoded
on its definition.
%
This approach has the advantage of being extendable with different sources of
structured information.
%
We proceed to introduce two extensions that help to address the problematic
testing scenarios presented in the previous section.

@@

We can follow a similar reasoning as before to represent the pattern matching
structure from a given function.
%
Consider the function |foo| defined in the previous section.
%
We can derive a new data type for each one of its pattern matchings, whose
fields represent each pattern variable present on the given pattern, abstracting
away every pattern variable of type |Exp|.
%
Concretely, we define the following data types:

@@

The first pattern of |foo| conteins two pattern variables (|x| and |y|) of type
|Exp| that are abstracted in the fields of |Mk_foo_1|.
%
Similarly, the second pattern of |foo| contains the pattern variable |x| of type
|Int| represented by the first field of |Mk_foo_2|, along with the pattern
variable |y| of type |Exp| that also is abstracted in the second field of
|Mk_foo_2|.

To evaluate these representations back to our target data type, we simply expand
them it into the concrete value represented by the original pattern:
%
\begin{align*}
  &\evalrep{|Mk_foo_1|\ x\ y}{Exp} = \\
  &\ \ |Add (Add eval_x_Exp (Val 50)) (Add (Val 25) eval_y_Exp)|\\
  &\evalrep{|Mk_foo_2|\ x\ y}{Exp} = \\
  &\ \ |Mul (Val 50) (Mul (Val x) eval_y_Exp)|
\end{align*}

And the random generation of pattern |HRep|s is defined in the same way as we
did before with data constructors |HRep|s:

@@

Finally, we can join the pattern matching representation of each clause of |foo|
into a single one:

@@

To introduce the higher level representation of module abstract interfaces,
suppose we have a module |M| defining |Exp| combinators as follows:

@@

We will represent each function of |M| returning a value of type |Exp| following
the same idea as before, deriving a data type with a single data constructor for
each one of them:

@@

In this case, each single constructor will have as fields the types of the
inputs of the function that they represent.
%
As before, we abstract away any field representing and input of type |Exp| with
a type parameter |r|.

To evaluate these representations, we simply pipe the values on its fields as
input parameter of each original function, returning its result.

@@

Moreover, the generation procedure for abstract interface representations
follows the same pattern as before:

@@

We will also define a type synonym to join all the representations of the module
|M| into a single one.

@@

Finally, we can put all the derived machinery together into a generation
specification |Spec| for the type |Exp|, assigning (possibly) different
generation frequencies to each individual |HRep| that constitutes it:

@@

This previous definition can be interpreted graphically as it is shown in the
Figure \ref{fig:hrep}, where curly arrows represent the structural information
extracted using meta-programming.

@@

This representation can be extended to mutually recursive and parametric types
...
